---
title: "Beyond maps: Patterns of formation processes at the Middle Pleistocene open-air site of Marathousa~1, Megalopolis Basin, Greece"
subtitle: "Point pattern analysis 3D"
author: "Domenico Giusti"
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: "../doc/manuscript/marathousa_rr.bib"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages: spatial analysis
library(sp)
library(rgdal)
library(raster)
library(maptools)
library(gtools)
library(mapproj)
library(spatstat)
```

## Spatial data collection and sampling

> The spatial 'window' containing the point pattern is currently required to be a 3D rectangual box [@Baddeley2016]

We restricted the three-dimensional point pattern analysis to a 5x4x0.7 cube meters region in area A, for which we have an unbiased database.

```{r}
getwd()
setwd("/home/dncgst/projects/marathousa1_qi")

# Area A (2015)
## bounding box, min/max vertical extent of lithics from UA3c, UA4
summary(Xa.ppp[Xa.ppp$marks$geo_takis %in% c("UA3c","UA4") & Xa.ppp$marks$class == "Lithic"])
b3 <- box3(c(938,943),c(673,677),c(349.1,349.9), unitname="meter")
b2 <- owin(xrange=c(938,943), yrange=c(673,677), unitname="meter")
## Elephas
Eleph.pp3 <- Elephas.ppp
summary(Eleph.pp3)
Eleph.pp3 <- with(Eleph.pp3, pp3(x,y,z,b3))
summary(Eleph.pp3)
## Lithic
summary(Xa.ppp)
lithic.A.pp3 <- Xa.ppp[Xa.ppp$marks$geo_takis %in% c("UA3c","UA4") & Xa.ppp$marks$class == "Lithic"]
summary(lithic.A.pp3)
lithic.A.pp3 <- as.data.frame(lithic.A.pp3)
lithic.A.pp3 <- with(lithic.A.pp3, pp3(x,y,z,b3))
summary(lithic.A.pp3)
plot(lithic.A.pp3)
## Fauna
summary(Xa.ppp)
fauna.A.pp3 <- Xa.ppp[Xa.ppp$marks$geo_takis %in% c("UA3c","UA4") & Xa.ppp$marks$class == "Fauna"]
fauna.A.pp3 <- as.data.frame(fauna.A.pp3)
fauna.A.pp3 <- with(fauna.A.pp3, pp3(x,y,z,b3))
plot(fauna.A.pp3)
## plot
plot(Eleph.pp3); plot(lithic.A.pp3, col="red", add=TRUE)
```

## Exploratory data analysis

Lithic:
Three-dimensional point pattern
54 points
Box: [938, 943] x [673, 677] x [349.1, 349.9] meter
Volume 15.9999999999991 cubic meter
Average intensity 3.37500000000019 points per cubic meter

Fauna:
Three-dimensional point pattern
286 points
Box: [938, 943] x [673, 677] x [349.1, 349.9] meter
Volume 15.9999999999991 cubic meter
Average intensity 17.875000000001 points per cubic meter

### Intensity

### Correlation

```{r}
# 3dimensional K-function
plot(K3est(lithic.A.pp3))
E <- envelope.pp3(lithic.A.pp3, fun=K3est, nsim=19, correction="iso", global=FALSE)
plot(E)

# 3dimensional pair corralation function
plot(pcf3est(lithic.A.pp3))
E <- envelope.pp3(lithic.A.pp3, fun=pcf3est, nsim=19, correction="iso", global=FALSE)
plot(E)
```

### Spacing

```{r}
# 3dimensional nearest-neighbour G function
plot(envelope.pp3(lithic.A.pp3, fun=G3est, nsim=19))
```

```{r}
# 3dimensional emty-space F function
plot(envelope.pp3(lithic.A.pp3, fun=F3est, nsim=19))
```

```{r, eval=FALSE, echo=FALSE}
# Computes the distances between pairs of points taken from two different three-dimensional point patterns
## d = \sqrt{(x_1-x_0)^2 + (y_1-y_0)^2 + (z_1 - z_0)^2}
## crossdist.pp3
## This function expects two point patterns X and Y, and returns the matrix whose [i,j] entry is the distance from X[i] to Y[j].
## nncross.pp3
## Given two point patterns X and Y in three dimensions, this function finds, for each point of X, the nearest point of Y. The distance between these points is also computed. If the argument k is specified, then the k-th nearest neighbours will be found. The return value is a data frame, with rows corresponding to the points of X. The first column gives the nearest neighbour distances (i.e. the ith entry is the distance from the ith point of X to the nearest element of Y). The second column gives the indices of the nearest neighbours (i.e.\ the ith entry is the index of the nearest element in Y.) If what="dist" then only the vector of distances is returned. If what="which" then only the vector of indices is returned. Note that this function is not symmetric in X and Y. To find the nearest neighbour in X of each point in Y, use nncross(Y,X).

nn <- nncross(lithic.A.pp3, Eleph.pp3, what="dist")
nn <- nncross(Eleph.pp3, lithic.A.pp3, what="dist")
summary(nn)
hist(nn, prob=TRUE, xlab="meter", main="b")
lines(density(nn, bw="nrd"), col="#a51e37", lwd=2)
```

## Fig.9

```{r}
# fig.: area A
setEPS()

#postscript("./doc/artwork/Fig9.eps", width=30, height=10)

pdf("./doc/artwork/Fig9.pdf", width=30, height=10)

par(mfcol=c(1,3), ps=12, cex=1, cex.main=2, cex.axis=1, cex.lab=1)

set.seed(0979)

w <- as.owin(unitsA_2015.shp)
plot(density(unmark(UA3c.ppp), sigma=bw.diggle(unmark(UA3c.ppp)), diggle=TRUE), axes=TRUE, main="a"); plot(w, add=TRUE); plot(b2, lwd=4, border="green", add=TRUE)
d <- density(UA3c.ppp[UA3c.ppp$marks$class == "Lithic"], sigma=bw.diggle(UA3c.ppp[UA3c.ppp$marks$class == "Lithic"]), diggle=TRUE)
contour.im(d, add=TRUE) # nlevels=4
e <- as.owin(Elephas.shp)
plot(e, add=TRUE, pch=24, lwd=2, col="grey") #, border="red")
GISTools::north.arrow(xb=937.5, yb=677.5, len=0.05, lab="N", cex.lab=1)  
GISTools::map.scale(xc=938.5, yc=677.5, len=1, units="meter", ndivs=1)


E <- envelope.pp3(lithic.A.pp3, fun=pcf3est, nsim=199, correction="iso", global=FALSE)
plot(E, xlim=c(0,1), lwd=2, legend=TRUE, main="b")
abline(v=0.35, lty=2, lwd=1, col="grey"); abline(v=0.65, lty=2, lwd=1, col="grey")
text(0.37,0, label="0.35", col="grey"); text(0.67,0, label="0.65", col="grey")

nn <- nncross(lithic.A.pp3, Eleph.pp3, what="dist")
hist(nn, prob=TRUE, xlab="meter", main="c")
lines(density(nn, bw="nrd"), col="#a51e37", lwd=2)

dev.off()
```

## References