---
title: "Beyond maps: Patterns of formation processes at the Middle Pleistocene open-air site of Marathousa~1, Megalopolis Basin, Greece"
subtitle: "Fabric analysis"
author: "Domenico Giusti"
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: "../doc/manuscript/marathousa_rr.bib"
output: 
  html_document: 
    fig_caption: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages: fabric analysis
library(circular)
library(CircStats)
library(RFOC)
library(plotrix)
library(heR.Misc)
```

## Reading and subsetting data

```{r}
getwd()
setwd("/home/dncgst/projects/marathousa1_qi")

# read the most updated db
edm_Kdb <- read.csv("data/edm_Kdb.csv", header=TRUE, sep=",", skip=0)

# duplicate orientation data (clock method)
edm_Kdb$dip_degree <- as.numeric(edm_Kdb$dip_clock)
edm_Kdb$orient_degree <- as.numeric(edm_Kdb$orient_clock)
class(edm_Kdb$dip_degree)
class(edm_Kdb$orient_degree)
summary(edm_Kdb)

# convert the clock system to degree
library(plyr)
edm_Kdb$dip_degree <- mapvalues(edm_Kdb$dip_degree, from=c(1,2,3,4,5,6), to=c(0,22,45,67,90,90))
edm_Kdb$orient_degree <- mapvalues(edm_Kdb$orient_degree, from=c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18), to=c(30,60,90,120,150,180,210,240,270,300,330,360,30,60,90,120,150,360))

# correct orient_degree! orient (clock) is recorded with reference to the excavation grid (-14°)
edm_Kdb$orient_degree <- edm_Kdb$orient_degree -14

# save
write.csv(edm_Kdb, file="data/edm_Kdb.csv", row.names=TRUE, na="NA")
```

```{r}
# subset coordinated materials:
finds <- c("LI","LOST.Lithic","MP","KN","LOST.KN","KN flake","ZN","MF","MO","OR","Eggshell","Flower","Fruit","HZ","Charcoal","Rock","Concretion","Limonite")

## Area A, excluding the Elephas (skeleton)
A <- edm_Kdb[which(edm_Kdb$edm_best %in% finds & edm_Kdb$edm_suffix == 0 & edm_Kdb$area == "A"),]
summary(A)

## Elephas (skeleton)
### read GK database
library(readxl)
E.skeleton <- read_xlsx("~/Dropbox/MAR-1_database/FOSSIL_LIST_MAR.xlsx", sheet = "Area A")
names(E.skeleton)
### subset Elephas (skeleton), excluding ?Elephas (skeleton)
E.skeleton <- E.skeleton[which(E.skeleton$GENUS == "Elephas (skeleton)"),c(2,3,4,16:22)]
### subset Elephas (skeleton) from 2013 to 2016
E.skeleton <- E.skeleton[-which(E.skeleton$YEAR == "2017"),]
### merge with total station db
E.skeleton <- merge(E.skeleton, edm_Kdb[which(edm_Kdb$edm_suffix == 0),], by=c("unit","edm_id"), all.x=TRUE)
summary(E.skeleton)

## Area B
B <- edm_Kdb[which(edm_Kdb$edm_best %in% finds & edm_Kdb$edm_suffix == 0 & edm_Kdb$area == "B"),]
```

```{r}
# subset material by units
fabric <- subset(A, geo_takis == "UA3c")
fabric <- subset(A, geo_takis == "UA4")
fabric <- E.skeleton
fabric <- subset(B, geo_takis == "UB4c")
fabric <- subset(B, geo_takis == "UB5a")
summary(fabric)

# duplicate dip_degree/orient_degree columns in plunge/bearing
fabric$plunge <- fabric[,"dip_degree"]
fabric$bearing <- fabric[,"orient_degree"]
is.numeric(fabric$plunge)
is.numeric(fabric$bearing)

# duplicate dip_comp/orient_comp columns in plunge/bearing (UB4c/UB5a compass)
fabric$plunge <- fabric[,"dip_comp"]
fabric$bearing <- fabric[,"orient_comp"]
is.numeric(fabric$plunge)
is.numeric(fabric$bearing)
## tidy plunge_comp
### length(grep(pattern=" [A-Z]+$", x=fabric$plunge))
fabric$plunge <- gsub(pattern=" [A-Z]+$", replacement="", x=fabric$plunge, ignore.case=FALSE)
# as numeric...
fabric$plunge <- as.numeric(fabric$plunge)

# remove NAs
summary(fabric)
fabric <- fabric[-which(is.na(fabric$bearing)),]

# subset elongated elements (NOT for the elephant)
## subset elements > 2cm 
fabric <- fabric[which(fabric$length >= 20),]
## subset elements with elengation index >= 1.6
fabric$elong_index <- fabric$length/fabric$width
fabric <- fabric[which(fabric$elong_index >= 1.6),]
## subset Elephas skeletal elements
fabric <- fabric[which(fabric$edm_best == "Elephant bone"),] # exclude Elephant skull, teeth and tusks
fabric$SKELETAL.ELEMENT <- as.factor(fabric$SKELETAL.ELEMENT)
#fabric <- fabric[-grep(pattern="vertebra", x=fabric$SKELETAL.ELEMENT),]
#fabric <- fabric[-grep(pattern="atlas", x=fabric$SKELETAL.ELEMENT),]
#fabric <- fabric[-grep(pattern="axis", x=fabric$SKELETAL.ELEMENT),]
#fabric <- fabric[-grep(pattern="phalanx", x=fabric$SKELETAL.ELEMENT),]
fabric <- fabric[-grep(pattern="astragalus", x=fabric$SKELETAL.ELEMENT),]
fabric <- fabric[-grep(pattern="rotule", x=fabric$SKELETAL.ELEMENT),]
fabric <- fabric[-grep(pattern="intermedium", x=fabric$SKELETAL.ELEMENT),]
## remove MF from UA3c
fabric <- fabric[-which(fabric$edm_best == "MF"),]
## remove MO from UB4c
fabric <- fabric[-which(fabric$edm_best == "MO"),]

# summary
summary(fabric)

# save and export
UA3c.fabric <- fabric
write.csv(UA3c.fabric, file="../marathousa1_qi/data/UA3c.fabric.csv", row.names=TRUE, na="NA")
UA4.fabric <- fabric
write.csv(UA4.fabric, file="../marathousa1_qi/data/UA4.fabric.csv", row.names=TRUE, na="NA")
Elephas.fabric <- fabric
write.csv(Elephas.fabric, file="../marathousa1_qi/data/Elephas.fabric.csv", row.names=TRUE, na="NA")
UB4c_clock.fabric <- fabric
write.csv(UB4c_clock.fabric, file="../marathousa1_qi/data/UB4c_clock.fabric.csv", row.names=TRUE, na="NA")
UB4c_comp.fabric <- fabric
write.csv(UB4c_comp.fabric, file="../marathousa1_qi/data/UB4c_comp.fabric.csv", row.names=TRUE, na="NA")
UB5a_clock.fabric <- fabric
write.csv(UB5a_clock.fabric, file="../marathousa1_qi/data/UB5a_clock.fabric.csv", row.names=TRUE, na="NA")
UB5a_comp.fabric <- fabric
write.csv(UB5a_comp.fabric, file="../marathousa1_qi/data/UB5a_comp.fabric.csv", row.names=TRUE, na="NA")
```

```{r}
# reproducible research from here...
getwd()
setwd("/home/dncgst/projects/marathousa1_qi")
UA3c.fabric <- read.csv("data/UA3c.fabric.csv", header=TRUE, sep=",", skip=0)
UA4.fabric <- read.csv("data/UA4.fabric.csv", header=TRUE, sep=",", skip=0)
Elephas.fabric <- read.csv("data/Elephas.fabric.csv", header=TRUE, sep=",", skip=0)
UB4c_clock.fabric <- read.csv("data/UB4c_clock.fabric.csv", header=TRUE, sep=",", skip=0)
UB4c_comp.fabric <- read.csv("data/UB4c_comp.fabric.csv", header=TRUE, sep=",", skip=0)
UB5a_clock.fabric <- read.csv("data/UB5a_clock.fabric.csv", header=TRUE, sep=",", skip=0)
UB5a_comp.fabric <- read.csv("data/UB5a_comp.fabric.csv", header=TRUE, sep=",", skip=0)
```

The subset of elongated materials for the fabric analysis includes:

Area A

In area A, most of the data regarding the orientation of the coordinated finds have been recorded prefering the clock system over the compass.

UA3c -- 49 obs (25HZ + 18KN + 1LI + 5ZN) with bearing/plunge information; elongation index >= 1.6, length >= 20mm

UA4 -- 38 obs (30HZ + 8KN) with bearing/plunge information; elongation index >= 1.6, length >= 20mm

Elephas (skeleton) -- 43 obs with bearing/plunge information (4 limb bones; pelvis (2 parts); 29 complete and rib frags; 2 calcaneus; 4 metatarsals/metacarpals; 1 pyramidal; 1 trapezoid)

Elephas (skeleton) -- 63 obs with bearing/plunge information (4 limb bones; pelvis (2 parts); 29 complete and rib frags; 2 calcaneus; 4 metatarsals/metacarpals; 1 pyramidal; 1 trapezoid; 1 atlas; 1 axis; 2 phalanx; 16 vertebra)

Area B

UB4c_clock -- 86 obs (61KN, 14LI, 3LOST.KN, 4ZN, 4HZ) with bearing/plunge information; elongation index >= 1.6, length >= 20mm

UB4c_comp -- 65 obs (43KN, 8LI, 10HZ, 2ZN, 2LOST.KN) with bearing/plunge information; elongation index >= 1.6, length >= 20mm

UB5a_clock -- 7 obs (6KN, 1LOST.KN) with bearing/plunge information; elongation index >= 1.6, length >= 20mm

UB5a_comp -- 7 obs (7KN) with bearing/plunge information; elongation index >= 1.6, length >= 20mm

## Global Circular statistics

### Rose diagrams (bearing)

#### Fig.4

```{r}
# fig.4: rose diagram, 12 bins (30 degrees)

setEPS()

#postscript("./doc/artwork/Fig4.eps", width=25, height=5)

pdf("./doc/artwork/Fig4_.pdf", width=25, height=5)

par(mfcol=c(1,5), ps=12, cex=1, cex.main=2, cex.axis=1, cex.lab=1)

heR.Misc::rose(UA3c.fabric$bearing, bins=12, col="#a51e37", rings=TRUE, labels=TRUE, rscale=2, cex=1, main="a")
heR.Misc::rose(UA4.fabric$bearing, bins=12, col="#a51e37", rings=TRUE, labels=TRUE, rscale=2, main="b")
heR.Misc::rose(Elephas.fabric$bearing, bins=12, col="#a51e37", rings=TRUE, labels=TRUE, rscale=2, main="c")
heR.Misc::rose(UB4c_clock.fabric$bearing, bins=12, col="#a51e37", rings=TRUE, labels=TRUE, rscale=2, main="d")
heR.Misc::rose(UB4c_comp.fabric$bearing, bins=12, col="#a51e37", rings=TRUE, labels=TRUE, rscale=2, main="e")

dev.off()
```

Area A

UA3c -- The rose diagram shows a unimodal distribution with prevalent peaks in the E range.

UA4 -- The rose diagram shows an unimodal distribution with prevalent peaks oriented N to E.

Elephant -- The rose diagram shows an unimodal distribution skewed to the South-West, with peaks in the NE quadrant.

Area B

UB4c_clock -- The rose diagram shows a multimodal distribution, with prevalent peaks oriented in the N and SE range.

UB4c_comp -- The rose diagram shows a unimodal distribution skewed to the West, with prevalent peaks oriented NE and SE.

### Two-dimensional summary statistics

```{r}
plunge <- rad(UA3c.fabric$plunge)
bearing <- rad(UA3c.fabric$bearing)
plunge <- rad(UA4.fabric$plunge)
bearing <- rad(UA4.fabric$bearing)
plunge <- rad(Elephas.fabric$plunge)
bearing <- rad(Elephas.fabric$bearing)
plunge <- rad(UB4c_clock.fabric$plunge)
bearing <- rad(UB4c_clock.fabric$bearing)
plunge <- rad(UB4c_comp.fabric$plunge)
bearing <- rad(UB4c_comp.fabric$bearing)
```

```{r}
# The mean angle (Azimuth) is calculated from sin and cos. The circular variance ranges from 0 to 1, where 0 means uniform dispersion and 1 concentration in one direction

## summary
deg(circ.summary(bearing))
deg(circ.summary(plunge))
## mean
deg(circ.mean(bearing))
deg(circ.mean(plunge))
## dispersion
deg(circ.disp(bearing))
deg(circ.disp(plunge))
```

Area A

UA3c -- Mean direction: 77.17° (E), circular variance: 41.95°; Mean dip: 20.35°, variance: 4.36°

UA4 -- Mean direction: 35.79° (N-NE), circular variance: 35.19°; Mean dip: 11.36°, variance: 2.15°

Elephas (skeleton) 43 obs -- Mean direction: 64.49° (NE), circular variance: 27.89°; Mean plunge: 10.68°, variance: 1.66°

Elephas (skeleton) 63 obs -- Mean direction: 54.64° (NE), circular variance: 29.25°; Mean plunge: 10.99°, variance: 3.13°

Area B

UB4c_clock -- Mean direction: 83.96° (E), circular variance: 47.72°; Mean dip: 36.14°, variance: 6.20°

UB4c_comp -- Mean direction: 128.14° (SE), circular variance: 44.41°; Mean dip: 25.26°, variance: 2.77°

To test the significance of the directional mean with Rayleigh z test: `r length(bearing)*(circ.disp(bearing)$var^2)` and compare with the critical z values table.

### Uniformity tests (bearing)

```{r}
# Isotropy (uniformity) tests
# As a general rule of thumb, for those cases where the null hypothesis is actually true, we do not want to incorrectly reject H0 (Type 1 error) more than 5% of the time. This correspond to a significance level of alpha=0.05 (95% confidence interval). Similarly, using a 99% confidence interval to evaluate a hypothesis is equivalent to a significance level of alpha=0.01.  The p-value is a way of quantifying he strength of evidence against the H0 and in favor of the alternative. We evaluate the hypotheses by comparing the p-value to the significance level: smaller p-value --> we reject H0!

## Rayleigh test
## test of uniformity, assessing the significance of the mean resultant length
### assumptions: data are not diametrically bidirectional; data are unimodal (there are no more than one cluster of points around the circle)
### uses the Vector Magnitude L parameter, that makes sense only if the distribution of orientations is not bi- or plurimodal
r.test(bearing)

## omnibus Kuiper test
## one-sample test of uniformity on the circle.
kuiper(bearing, alpha=0.01)

## omnibus Watson test
## goodness of fit test for the von Mises (dist="vm") or circular uniform (dist="uniform") distribution
watson(bearing, alpha=0.01, dist = "uniform")

## omnibus Rao's spacing test
## test of uniformity
rao.spacing(bearing, alpha=0.01)

## Rao's tests for homogeneity
## rao.homogeneity(as.list(bearing), alpha=0.01)
```

Area A

UA3c -- All the tests clearly reject the null hypothesis of uniformity, thus supporting an anisotropic distribution of the sample.

UA4 -- All the tests clearly support an anisotropic distribution of the sample.

Elephant -- Rayleigh test significantly reject the null hypothesis of uniformity in the orientation of the sample. Kuiper's Test of Uniformity do reject as well the null hypothesis. Watson's goodness of fit test for the uniform distribution do reject the null hypothesis. Rao's Spacing Test of Uniformity do reject the null hypothesis of uniformity. All the tests support a strong anisotropic distribution of the sample.

Elephas (63 obs.) -- Rayleigh: r.bar 0.4895319 p.value 2.775087e-07
                     Kuiper: 3.4811 (2.001)
                     Watson: 0.906 (0.267)
                     Rao: 291.4286 (155.49)

Area B

UB4c_clock -- Rayleigh and Watson tests do not reject the null hypothesis; Kuiper and Rao do reject the null hypothesis...

UB4c_comp -- Rayleigh test do reject the null hypothesis of uniformity in the orientation of the sample (with 95% confidence level). Kuiper's Test of Uniformity do not reject the null hypothesis. Watson's goodness of fit test for the uniform distribution do not reject the null hypothesis. Rao's Spacing Test of Uniformity do not reject the null hypothesis of uniformity, but the test statistic is very close to the critical value. All the tests, support an isotropic distribution of the sample.

### Equal area (Schmidt) diagrams

```{r}
# equal area (schmidt) diagram
net()
qpoint(deg(bearing), 90-deg(plunge), col=4, pch=16, cex=.5)
#lowplane(270, 3.908544, col=2, UP=TRUE, PLOT=TRUE)
```

### Randomness testing in three-dimensional orientation data

```{r}
plunge <- UA3c.fabric$plunge
bearing <- UA3c.fabric$bearing
plunge <- UA4.fabric$plunge
bearing <- UA4.fabric$bearing
plunge <- Elephas.fabric$plunge
bearing <- Elephas.fabric$bearing
plunge <- UB4c_clock.fabric$plunge
bearing <- UB4c_clock.fabric$bearing
plunge <- UB4c_comp.fabric$plunge
bearing <- UB4c_comp.fabric$bearing
```

> One problem in applying the S1/S3 test is that the eigenvector method cannot cope adeguately with certain types of multimodal data [@Woodcock1977]. For this reason the test, as with most other tests reviewed here, could wrongly return a random result from a dataset consisting of several individually significant modes or superimposed samples. [...] For this reason we recommended inspectiong the form of the data on an equal-area plot before accepting the results of any randomness test. If several modes appear to be present then some independent test for multimodality [e.g. @Bailey1975] should be made, together with some attempt to separate data from the various modes [@Kohlbeck&Scheidegger1977] before proceeding further [@Woodcock1983,p.548].

Woodcock [-@Woodcock1983] test based on eigenvalue method of orientation data analysis.

```{r}
# Stepwise procedure for the S1/S3 test. Modified from [@Woodcock1983]
## step 1) express each direction in the sample as a set of three direction cosines. If the data are in the form of trend and plunge, the formulae are:
### l=cos(plunge)*cos(trend)
l <- c(cos(plunge)*cos(bearing))
### m=cos(plunge)*sin(trend)
m <- c(cos(plunge)*sin(bearing))
### n=sin(plunge)
n <- c(sin(plunge))
## step 2) and 3) compute the products of each direction-cosine set. sum the products over the whole data sample to give the orientation tensor
b <- matrix(c(
  sum(l^2),sum(l*m),sum(l*n),
  sum(m*l),sum(m^2),sum(m*n),
  sum(n*l),sum(n*m),sum(n^2)),
  ncol=3, byrow=TRUE)
## step 4) compute the eigenvectors and eigenvalues of this matrix
### the eigenvectors (v1,v2,v3) of this matrix correspond to three orthogonal 'principal axes' of the data sample. Axis v1 is an estimate of the mean direction of the data; v3 is the pole to any girdle present in the data [@Watson1966]
eigenvectors <- eigen(b)$vectors
dimnames(eigenvectors) <- list(c("v1","v2","v3"),c("l","m","n"))
### The eigenvalues (λ1,λ2,λ3), associated with (v1,v2,v3) respectively, describe the shape of the data sample [@Watson1966; @Mark1974; @Woodcock1973; @Woodcock1977]
eigenvalues <- eigen(b)$values
names(eigenvalues) <- c("λ1","λ2","λ3")
```

```{r}
## equal area plot
v1.plunge <- deg(asin(eigenvectors["v1","n"]))
v1.bearing <- deg(acos(eigenvectors["v1","l"]/cos(asin(eigenvectors["v1","n"]))))
v2.plunge <- deg(asin(eigenvectors["v2","n"]))
v2.bearing <- deg(acos(eigenvectors["v2","l"]/cos(asin(eigenvectors["v2","n"]))))
v3.plunge <- deg(asin(eigenvectors["v3","n"]))
v3.bearing <- deg(acos(eigenvectors["v3","l"]/cos(asin(eigenvectors["v3","n"]))))
net()
qpoint(deg(bearing), 90-deg(plunge), col=1, pch=1, cex=1)
qpoint(deg(v1.bearing), 90-deg(v1.plunge), col=4, pch=10, cex=1, lab="v1")
qpoint(deg(v2.bearing), 90-deg(v2.plunge), col=4, pch=10, cex=1, lab="v2")
qpoint(deg(v3.bearing), 90-deg(v3.plunge), col=4, pch=10, cex=1, lab="v3")
#lowplane(270, 3.908544, col="red", UP=TRUE, PLOT=TRUE)
```

```{r}
## step 5) compute the eigenvalue ratio, particularly the ratio of the largest to smallest eigenvalue (S1/S3=λ1/λ3)
### since λ1+λ2+λ3=N then normalized eigenvalues Sj=λj/N are related by S1+S2+S3=1. A data cluster has S1>S2~=S3 and a data girdle has S1~=S2>S3
s1 <- eigenvalues["λ1"]/length(plunge)
s2 <- eigenvalues["λ2"]/length(plunge)
s3 <- eigenvalues["λ3"]/length(plunge)
```

```{r}
# save
## UA3c
UA3c.s1_s2 <- s1/s2
UA3c.s2_s3 <- s2/s3
UA3c.s1_s3 <- s1/s3
## UA4
UA4.s1_s2 <- s1/s2
UA4.s2_s3 <- s2/s3
UA4.s1_s3 <- s1/s3
# Elephant (43 obs)
E.s1_s2 <- s1/s2
E.s2_s3 <- s2/s3
E.s1_s3 <- s1/s3
# Elephant (63 obs)
E63.s1_s2 <- s1/s2
E63.s2_s3 <- s2/s3
E63.s1_s3 <- s1/s3
# UB4c_clock
UB4c_clock.s1_s2 <- s1/s2
UB4c_clock.s2_s3 <- s2/s3
UB4c_clock.s1_s3 <- s1/s3
# UB4c_comp
UB4c_comp.s1_s2 <- s1/s2
UB4c_comp.s2_s3 <- s2/s3
UB4c_comp.s1_s3 <- s1/s3
```

```{r}
## step 6) refer to Fig.9 or Table 1 and read off the critical S1/S3 value for the appropriate sample size and required confidence level. Note that for a small sample size a 95% confidence level is appropriate
### for N=300 the S1/S3 critical value at 95% confidence level is: 1.34
## step 7) if the computed value is _less_ than the test value, the data sample cannot be distinguished from a random sample at this confidence level
s1s3_test <- read.csv(url("https://raw.githubusercontent.com/dncgst/woodcock/master/vignettes/Woodcock%26Naylor1983_fig9.csv"), header=TRUE, sep=",", skip=0)
str(s1s3_test)
#C <- as.data.frame(cbind(c(UA3c.s1_s3, UA4.s1_s3, E.s1_s3, UB4c_clock.s1_s3, UB4c_comp.s1_s3),c(49, 38, 43, 86, 65)))
C <- as.data.frame(cbind(c(UA3c.s1_s3, UA4.s1_s3, E63.s1_s3, UB4c_clock.s1_s3, UB4c_comp.s1_s3),c(49, 38, 63, 86, 65)))
```

#### Fig.5

```{r}
# fig.5: Woodcock diagram

setEPS()

#postscript("./doc/artwork/Fig5.eps", width=20, height=10)

pdf("./doc/artwork/Fig5_.pdf", width=20, height=10)

par(mfcol=c(1,2), ps=12, cex=1, cex.main=2, cex.axis=1, cex.lab=1)

## eigenvalues ratio graph
plot(log(UA3c.s2_s3),log(UA3c.s1_s2),pch=15,xlim=c(0,3),ylim=c(0,3),xlab="ln S2/S3",ylab="ln S1/S2", main="a")
## shape parameter K=ln(S1/S2)/ln(S2/S3) ranges from 0 (uniaxial girdles) to infinite (uniaxial clusters)
abline(h=0,col="grey",lty=2)
abline(v=0,col="grey",lty=2)
abline(0,1,col="grey",lty=2)
abline(0,2,col="grey",lty=2)
abline(0,.5,col="grey",lty=2)
abline(0,5,col="grey",lty=2)
abline(0,.2,col="grey",lty=2)
text(2.95,2.8,label="K=1",col="grey",lwd=.02)
text(2.95,.12,label="K=0",col="grey",lwd=.02)
text(.2,3,label=expression(paste("K=",infinity)),col="grey",lwd=.02)
text(2.95,.75,label="K=0.2",col="grey",lwd=.02)
text(2.95,1.65,label="K=0.5",col="grey",lwd=.02)
text(.8,3,label="K=5",col="grey",lwd=.02)
text(1.7,3,label="K=2",col="grey",lwd=.02)
## complementary to the shape parameter is the parameter C=ln(S1/S3) which expresses the 'strength' of the preferred orientation in the data sample [@Woodcock1977]. Strongly organized samples have larger C and plot further from the origin of the ratio graph. A perfect uniform distribution of data would have C=0 and would plot at the origin with S1=S2=S3=1/3. Random samples will also plot close to the origin and have low C values, suggesting that this parameter might be used as a test statistic for randomness
abline(3,-1,col="grey",lty=1)
text(1.7,1.5,label="C=3",col="grey",lwd=.02)
##
PlotPTsmooth(UA3c.fabric$bearing,UA3c.fabric$plunge,x=log(UA3c.s2_s3),y=log(UA3c.s1_s2),siz=0.1,IMAGE=TRUE,CONT=TRUE,add=TRUE)
text(.8,.4,label="UA3c",col="black",lwd=.04)
PlotPTsmooth(UA4.fabric$bearing,UA4.fabric$plunge,x=log(UA4.s2_s3),y=log(UA4.s1_s2),siz=0.1,IMAGE=TRUE,CONT=TRUE,add=TRUE)
text(1.9,.6,label="UA4",col="black",lwd=.04)
#PlotPTsmooth(Elephas.fabric$bearing,Elephas.fabric$plunge,x=log(E.s2_s3),y=log(E.s1_s2),siz=0.1,IMAGE=TRUE,CONT=TRUE,add=TRUE)
#text(2.1,.1,label="Eleph.",col="black",lwd=.04)
PlotPTsmooth(Elephas.fabric$bearing,Elephas.fabric$plunge,x=log(E63.s2_s3),y=log(E63.s1_s2),siz=0.1,IMAGE=TRUE,CONT=TRUE,add=TRUE)
text(1.8,.1,label="Eleph.",col="black",lwd=.04)
PlotPTsmooth(UB4c_clock.fabric$bearing,UB4c_clock.fabric$plunge,x=log(UB4c_clock.s2_s3),y=log(UB4c_clock.s1_s2),siz=0.1,IMAGE=TRUE,CONT=TRUE,add=TRUE)
text(.4,.02,label="UB4c (clock)",col="black",lwd=.04)
PlotPTsmooth(UB4c_comp.fabric$bearing,UB4c_comp.fabric$plunge,x=log(UB4c_comp.s2_s3),y=log(UB4c_comp.s1_s2),siz=0.1,IMAGE=TRUE,CONT=TRUE,add=TRUE)
text(.5,.5,label="UB4c (comp)",col="black",lwd=.04)

##
text(1.45,1.65,label="WEAK <--> STRONG",col="gray",lwd=.03,srt=45,pos=1)
text(2.55,2.63,label="<-- CLUSTERS | GIRDLES -->",col="gray",lwd=.03,srt=-45,pos=1)

### plot
plot(x=s1s3_test$N, y=log(s1s3_test$X90.), type="b", col="grey", lty=2, xlim=c(0,100), xlab="Obs.", ylab="ln C", main="b")
lines(x=s1s3_test$N, y=log(s1s3_test$X95.), type="b", col="grey", lty=2, add=TRUE)
lines(x=s1s3_test$N, y=log(s1s3_test$X97.5.), type="b", col="grey", lty=2, add=TRUE)
lines(x=s1s3_test$N, y=log(s1s3_test$X99.), type="b", col="grey", lty=2, add=TRUE)
points(C$V2, log(C$V1), pch=21, bg="#a51e37", col="#a51e37", cex=2, add=TRUE)
text(10,1, label="RANDOM",col="grey", lwd=.02)
text(25,2, label="SIGNIFICANT",col="grey",lwd=.02)
text(49,1.05, label="UA3c",col="black",lwd=.02)
text(38,2.45, label="UA4",col="black", lwd=.02)
text(63,2.2, label="Eleph.",col="black", lwd=.02)
text(86,.4, label="UB4c (clock)",col="black", lwd=.02)
text(66,.95, label="UB4c (comp)",col="black", lwd=.02)

dev.off()
```

UA3c -- (49 obs.) The s1/s3 value: 2.535006 (`r UA3c.s1_s3`) is slightly higher than the critical S1/S3 test value (2.073) for N=50 at 95% confidence, thus the data sample reject the hypothesis of randomness in favor of a strong organized sample. The K value: 0.6074483 (`r log(UA3c.s1_s2)/log(UA3c.s2_s3)`) plots the data sample close to K=0.5, indicating uniaxial girdles

UA4 -- (38 obs.) The s1/s3 value: 10.4194 (`r UA4.s1_s3`) is definitely higher than the critical S1/S3 test value (2.28) for N=40 at 95% confidence, thus the data sample reject the hypothesis of randomness in favor of a strong organized sample. The K value: 0.2312978 (`r log(UA4.s1_s2)/log(UA4.s2_s3)`) plots the data sample close to K=0.2, indicating uniaxial girdles

Elephant -- (43 obs.) The s1/s3 value: 10.85199 (`r E.s1_s3`) is definitely higher than the critical S1/S3 test value (2.28) for N=40 at 95% confidence, thus the data sample reject the hypothesis of randomness in favor of a strong organized sample. The K value: 0.1401002 (`r log(E.s1_s2)/log(E.s2_s3)`) plots the data sample close to K=0.1, indicating uniaxial girdles.

Elephant -- (63 obs.)

UB4c_clock -- (86 obs.) The C value: 1.71685 (`r UB4c_clock.s1_s3`) is slightly higher than, yet very close to, the critical S1/S3 test value (1.70) for N=90 at 95% confidence, thus the data sample cannot reject the hypothesis of randomness in favor of a strong organized sample. The K value: 0.506163 (`r log(UB4c_clock.s1_s2)/log(UB4c_clock.s2_s3)`) plots the data sample between K=0 and K=1, indicating weak uniaxial girdles.

UB4c_comp -- (65 obs.) The C value: 2.309044 (`r UB4c_comp.s1_s3`) is higher than the critical S1/S3 test value (1.83) for N=70 at 95% confidence, thus the data sample reject the hypothesis of randomness in favor of a weak organized sample. The K value: 0.7039872 (`r log(UB4c_comp.s1_s2)/log(UB4c_comp.s2_s3)`) plots the data sample close to K=1.

### Benn diagram

```{r}
# isotropy index
UA3c.is <- s3/s1
UA4.is <- s3/s1
E.is <- s3/s1
E63.is <- s3/s1
UB4c_clock.is <- s3/s1
UB4c_comp.is <- s3/s1

# elongation index
UA3c.el <- 1-(s2/s1)
UA4.el <- 1-(s2/s1)
E.el <- 1-(s2/s1)
E63.el <- 1-(s2/s1)
UB4c_clock.el <- 1-(s2/s1)
UB4c_comp.el <- 1-(s2/s1)

# 3rd axis
UA3c.bs <- 1-(UA3c.is+UA3c.el)
UA4.bs <- 1-(UA4.is+UA4.el)
E.bs <- 1-(E.is+E.el)
E63.bs <- 1-(E63.is+E63.el)
UB4c_clock.bs <- 1-(UB4c_clock.is+UB4c_clock.el)
UB4c_comp.bs <- 1-(UB4c_comp.is+UB4c_comp.el)
```

#### Fig.6

```{r}
# fig.6: Benn diagram

setEPS()

#postscript("./doc/artwork/Fig6.eps", width=12, height=10)

pdf("./doc/artwork/Fig6_.pdf", width=12, height=10)

par(ps=12, cex=1, cex.main=2, cex.axis=1, cex.lab=1)

# benn diagram
triax.plot(matrix(c(UA3c.bs,UA3c.el,UA3c.is),ncol=3,byrow=2), show.grid=TRUE, pch=16, cex.ticks=.8, cex.axis=1, axis.labels=c("","Elongation Index\n EL=1-(S2/S1)","Isotropy Index\n IS=S3/S1"), tick.labels=list(l=seq(0.1,0.9,by=0.1), r=seq(0.1,0.9,by=0.1), b=NULL)) #, main="Benn diagram")
triax.points(matrix(c(UA3c.bs,UA3c.el,UA3c.is),ncol=3,byrow=2), pch=16, col="#a51e37", cex=2)
text(.44,.33, label="UA3c", lwd=.02)
triax.points(matrix(c(UA4.bs,UA4.el,UA4.is),ncol=3,byrow=2), pch=16, col="#a51e37", cex=2)
text(.4,.11, label="UA4", lwd=.02)
#triax.points(matrix(c(E.bs,E.el,E.is),ncol=3,byrow=2), pch=16, col="#a51e37", cex=2)
#text(.3,.11, label="Eleph.", lwd=.02)
triax.points(matrix(c(E63.bs,E63.el,E63.is),ncol=3,byrow=2), pch=16, col="#a51e37", cex=2)
text(.3,.14, label="Eleph.", lwd=.02)
triax.points(matrix(c(UB4c_clock.bs,UB4c_clock.el,UB4c_clock.is),ncol=3,byrow=2), pch=16, col="#a51e37", cex=2)
text(.42,.53, label="UB4c (clock)", lwd=.02)
triax.points(matrix(c(UB4c_comp.bs,UB4c_comp.el,UB4c_comp.is),ncol=3,byrow=2), pch=16, col="#a51e37", cex=2)
text(.48,.4, label="UB4c (comp)", lwd=.02)
text(0.5,0.9, label="isotropic")
text(-0.1,0, label="planar")
text(1.1,0, label="linear")

dev.off()
```

## References