---
title: "Beyond maps: Patterns of formation processes at the Middle Pleistocene open-air site of Marathousa~1, Megalopolis Basin, Greece"
subtitle: "Vertical distribution analysis"
author: "Domenico Giusti"
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: "../doc/manuscript/marathousa_rr.bib"
output: 
  html_document: 
    fig_caption: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages: spatial analysis
library(sp)
library(rgdal)
library(raster)
library(maptools)
library(gtools)
library(mapproj)
library(spatstat)
```

## Reading and subsetting data

```{r}
# read data
getwd()
setwd("/home/dncgst/projects/marathousa1_qi")

lithic_Kdb <- read.csv("data/lithic_Kdb.csv", header=TRUE, sep=",", skip=0)

## Area A
### subset archaeological and paleontological material from area A
lithic.A <- lithic_Kdb[which(lithic_Kdb$area == "A"),]
fossil_KdbA <- read.csv("data/fossil_KdbA.csv", header=TRUE, sep=",", skip=0)
fossil.A <- fossil_KdbA[-which(fossil_KdbA$ORDER == "Proboscidea"),] # subset fossils not attribuited to order Proboscidea. This will exclude also tusk fragments which are probably, yet not certainly, attribuited to the same individual
names(lithic.A)
foo <- lithic.A[ , c("x","y","z","edm_best","Lithic.class","geo_takis","spit","length","width","thickness","unit","edm_id","date")]
levels(foo$Lithic.class)
names(foo)[5] <- "type"
names(fossil.A)
bar <- fossil.A[ , c("x","y","z","edm_best","TYPE","geo_takis","spit","length","width","thickness","unit","edm_id","date")]
levels(bar$TYPE)
names(bar)[5] <- "type"
Xa <- rbind(foo, bar)
summary(Xa)
Xa$type <- factor(Xa$type)
Xa$class <- Xa$type
levels(Xa$class) <- c("Lithic","Lithic","Lithic","Lithic","Fauna","Fauna","Fauna","Fauna")
summary(Xa)
Xa <- Xa[-which(is.na(Xa$x)),]

### jitter coordinates for sieve finds

set.seed(0979)

Xa$edm_best <- factor(Xa$edm_best)

Xa$x[which(Xa$edm_best %in% c("Sample FLO","SF","SFA","SFB","SFPCN","SFPCW"))] <- jitter(Xa$x[which(Xa$edm_best %in% c("Sample FLO","SF","SFA","SFB","SFPCN","SFPCW"))], 1, 0.5)

Xa$y[which(Xa$edm_best %in% c("Sample FLO","SF","SFA","SFB","SFPCN","SFPCW"))] <- jitter(Xa$y[which(Xa$edm_best %in% c("Sample FLO","SF","SFA","SFB","SFPCN","SFPCW"))], 1, 0.5)

Xa$z[which(Xa$edm_best %in% c("Sample FLO","SF","SFA","SFB","SFPCN","SFPCW"))] <- Xa$z[which(Xa$edm_best %in% c("Sample FLO","SF","SFA","SFB","SFPCN","SFPCW"))] + 0.05
Xa$z[which(Xa$edm_best %in% c("Sample FLO","SF","SFA","SFB","SFPCN","SFPCW"))] <- jitter(Xa$z[which(Xa$edm_best %in% c("Sample FLO","SF","SFA","SFB","SFPCN","SFPCW"))], 1, 0.05)
summary(Xa)

### save and export
write.csv(Xa, file="data/Xa.csv", row.names=TRUE, na="NA")

## Area B
### subset archaeological and paleontological material from area B
lithic.B <- lithic_Kdb[which(lithic_Kdb$area == "B"),]
names(lithic.B)
foo <- lithic.B[ , c("x","y","z","edm_best","Lithic.class","geo_takis","spit","length","width","thickness","unit","edm_id","date")]
levels(foo$Lithic.class)
names(foo)[5] <- "type"
fossil_KdbB <- read.csv("data/fossil_KdbB.csv", header=TRUE, sep=",", skip=0)
fossil.B <- fossil_KdbB
names(fossil.B)
bar <- fossil.B[ , c("x","y","z","edm_best","TYPE","geo_takis","spit","length","width","thickness","unit","edm_id","date")]
names(bar)[5] <- "type"
Xb <- rbind(foo, bar)
summary(Xb)
Xb$type <- factor(Xb$type)
Xb$class <- Xb$type
levels(Xb$class) <- c("Bone technology","Lithic","Lithic","Lithic","Lithic","Lithic", "Fauna","Bone technology","Fauna","Fauna")
Xb <- Xb[-which(is.na(Xb$x)),]
## remove double entries: bone technology & KN flake
Xb[which(Xb$class == "Bone technology"),]
Xb <- Xb[-which(Xb$type == "KN flake"),]

### jitter collected finds coordinates
Xb$edm_best <- factor(Xb$edm_best)

Xb$x[which(Xb$edm_best %in% c("Sample FLO","SF","SFA","SFAC","SFB","SFC","SFD"))] <- jitter(Xb$x[which(Xb$edm_best %in% c("Sample FLO","SF","SFA","SFAC","SFB","SFC","SFD"))], 1, 0.25)

Xb$y[which(Xb$edm_best %in% c("Sample FLO","SF","SFA","SFAC","SFB","SFC","SFD"))] <- jitter(Xb$y[which(Xb$edm_best %in% c("Sample FLO","SF","SFA","SFAC","SFB","SFC","SFD"))], 1, 0.25)

Xb$z[which(Xb$edm_best %in% c("Sample FLO","SF","SFA","SFAC","SFB","SFC","SFD"))] <- Xb$z[which(Xb$edm_best %in% c("Sample FLO","SF","SFA","SFAC","SFB","SFC","SFD"))] + 0.025
Xb$z[which(Xb$edm_best %in% c("Sample FLO","SF","SFA","SFAC","SFB","SFC","SFD"))] <- jitter(Xb$z[which(Xb$edm_best %in% c("Sample FLO","SF","SFA","SFAC","SFB","SFC","SFD"))], 1, 0.025)
summary(Xb)

#### save and export
write.csv(Xb, file="data/Xb.csv", row.names=TRUE, na="NA")

## Elephas
### subset elephant coordinates, until 2015
Elephas.ppp <- edm_Kdb[which(edm_Kdb$edm_best %in% c("Elephant bone","Elephant skull","Elephant tooth","Elephant tusk")),c("unit","edm_id","edm_suffix","edm_best","x","y","z","date")]
class(Elephas.ppp$date)
Elephas.ppp$date <- as.Date(Elephas.ppp$date, format="%Y-%m-%d")
library(dplyr)
Elephas.ppp <- dplyr::filter(Elephas.ppp, date < as.Date("2016-01-01"))
write.csv(Elephas.ppp, file="../marathousa1_qi/data/Elephas.ppp.csv", row.names=TRUE, na="NA")
```

```{r}
# reproducible research from here...
getwd()
setwd("/home/dncgst/projects/marathousa1_qi")
Xa <- read.csv("data/Xa.csv", header=TRUE, sep=",", skip=0)
Xb <- read.csv("data/Xb.csv", header=TRUE, sep=",", skip=0)
Elephas.ppp <- read.csv("data/Elephas.ppp.csv", header=TRUE, sep=",", skip=0)
```

## IDW interpolation

```{r}
# read selection of surface points, Area A
UA4_surf <- read.csv("data/UA4_surf.csv", header=TRUE, sep=",", skip=0)

# window area A
ogrInfo("data/shp/AreaA_2015.shp")
unitsA_2015.shp <- readOGR("data/shp/AreaA_2015.shp", layer="AreaA_2015")
proj4string(unitsA_2015.shp) <- CRS(as.character(NA))
bbox(unitsA_2015.shp)

# convert to ppp
w <- as.owin(unitsA_2015.shp)
plot(w)
m <- as.numeric(UA4_surf$z)
UA4_surf <- ppp(UA4_surf$x, UA4_surf$y, window=w, marks=m)
UA4_surf <- as.ppp(UA4_surf)
any(duplicated(UA4_surf))
summary(UA4_surf)
plot(UA4_surf)

# IDW interpolation
UA4_surf.idw <- spatstat::idw(UA4_surf, power=10, at="pixels", eps=0.01)
plot(UA4_surf.idw, axes=TRUE, main="UA3c/4 erosional contact"); plot(w, add=TRUE)

# read selection of surface points, Area B
UB5a_surf <- read.csv("data/UB5a_surf_improved.csv", header=TRUE, sep=",", skip=0)

# windows area B
ogrInfo("data/shp/AreaB_2015.shp")
unitsB_2015.shp <- readOGR("data/shp/AreaB_2015.shp", layer="AreaB_2015")
proj4string(unitsB_2015.shp) <- CRS(as.character(NA))
bbox(unitsB_2015.shp)

# convert to ppp
w <- as.owin(unitsB_2015.shp)
m <- as.numeric(UB5a_surf$z)
UB5a_surf <- ppp(UB5a_surf$x, UB5a_surf$y, window=w, marks=m)
UB5a_surf <- as.ppp(UB5a_surf)
any(duplicated(UB5a_surf))
summary(UB5a_surf)
plot(UB5a_surf)

# IDW interpolation
UB5a_surf.idw <- spatstat::idw(UB5a_surf, power=10, at="pixels", eps=0.01)
plot(UB5a_surf.idw, axes=TRUE, main="UB4c/5a erosional contact"); plot(w, add=TRUE)
```

### Fig.3

```{r}
# fig.3: IDW interpolated UA3c/UA4 and UB4c/UB5a surfaces

setEPS()

postscript("./doc/artwork/Fig3.eps", width=10, height=10)

pdf("./doc/artwork/Fig3.pdf", width=30, height=10)

par(mfcol=c(1,3), ps=12, cex=1, cex.main=2, cex.axis=1, cex.lab=1)

w <- as.owin(unitsA_2015.shp)
plot(UA4_surf.idw, axes=TRUE, main="a"); plot(w, add=TRUE)
library(GISTools)
GISTools::north.arrow(xb=937.5, yb=677.5, len=0.05, lab="N", cex.lab=1)  
GISTools::map.scale(xc=938.5, yc=677.5, len=1, units="meter", ndivs=1)

w <- as.owin(unitsB_2015.shp)
plot(UB5a_surf.idw, axes=TRUE, main="b"); plot(w, add=TRUE)
GISTools::north.arrow(xb=933, yb=610, len=0.1, lab="N", cex.lab=1)
GISTools::map.scale(xc=933.5, yc=608.5, len=1, units="meter", ndivs=1)

z <- 2 * volcano        # Exaggerate the relief
x <- 10 * (1:nrow(z))   # 10 meter spacing (S to N)
y <- 10 * (1:ncol(z))   # 10 meter spacing (E to W)
persp(x, y, z, theta=135, phi=10, col="gray", scale=FALSE,
      ltheta=-120, shade=0.75, border=NA, box=FALSE, main="c")

dev.off()
```

## Calculate minimum distance

### Area A

```{r}
# subset points until 2015
summary(Xa)
Xa$date <- as.Date(Xa$date, format="%Y-%m-%d")
library(dplyr)
Xa <- dplyr::filter(Xa, date < as.Date("2016-01-01"))

# convert dataframe to ppp, Area A
w <- as.owin(unitsA_2015.shp)
names(Xa)
m <- Xa[,3:14]
Xa.ppp <- ppp(Xa$x, Xa$y, window=w, marks=m)
unitname(Xa.ppp) <- c("meter","meters")
summary(Xa.ppp)
## remove rejected points
Xa.ppp <- as.ppp(Xa.ppp)
any(duplicated(Xa.ppp))

## plot
plot(Xa.ppp$x, Xa.ppp$y, axes=TRUE, asp=1); plot(w, add=TRUE)
plot(Xa.ppp$marks$class, main="Distribution of remains from area A")
plot(droplevels(Xa.ppp$marks$geo_takis), main="Distribution of remains from area A")
```

Xa.ppp
Marked planar point pattern:  405 points
Average intensity 16.2 points per square meter
Window area = 25 square meters
UA3c     :279
UA4      : 61

```{r}
# calculate minimum distance?  Difficult to quantify due to its not sharp nature --> "Likewise, in Area A the relative vertical distribution of remains from UA3c was estimated relative to the absolute elevation of the elephant remains and the range of elevations of the UA3c/4 surface.""
summary(Xa.ppp)
S <- UA4_surf.idw[Xa.ppp]
summary(S)
d <- Xa.ppp$marks$z-S
Da <- cbind(as.data.frame(Xa.ppp), d)
summary(Da)
```

Da: 405 obs.

```{r}
Da <- as.data.frame(Xa.ppp)
```


#### Fig.7

```{r}
# fig.7: Vertical distribution area A

setEPS()

#postscript("./doc/artwork/Fig7.eps", width=10, height=10)

pdf("./doc/artwork/Fig7.pdf", width=10, height=10)

par(ps=12, cex=1, cex.main=2, cex.axis=1, cex.lab=1)

## UA3c
hist(0, prob=TRUE, xlim=c(348.5,350), ylim=c(0,4), xlab=" meters a.s.l.", main="")
rug(UA4_surf$marks, col="grey")
rect(xleft=quantile(UA4_surf$marks, .25), xright=quantile(UA4_surf$marks, .75), ybottom=0, ytop=7, col="grey", border=NA, lwd=2)
### Elephas
hist(Elephas.ppp$z, nclass=30, prob=TRUE, col=NA, add=TRUE)
### debris UA3c
rug(Da[which(Da$geo_takis == "UA3c" & Da$type == "debris"),]$z, col="#e41a1c", lwd=2)
lines(density(Da[which(Da$geo_takis == "UA3c" & Da$type == "debris"),]$z, bw="nrd"), col="#e41a1c", lwd=2)
abline(v=mean(Da[which(Da$geo_takis == "UA3c" & Da$type == "debris"),]$z), col="#e41a1c", lwd=1, lty=2)
### debitage UA3c
rug(Da[which(Da$geo_takis == "UA3c" & Da$type == "debitage"),]$z, col="#377eb8", lwd=2)
abline(v=mean(Da[which(Da$geo_takis == "UA3c" & Da$type == "debitage"),]$z), col="#377eb8", lwd=1, lty=2)
### tool UA3c
rug(Da[which(Da$geo_takis == "UA3c" & Da$type == "tool"),]$z, col="#4daf4a", lwd=2)
abline(v=mean(Da[which(Da$geo_takis == "UA3c" & Da$type == "tool"),]$z), col="#4daf4a", lwd=1, lty=2)
### fauna UA3c
lines(density(Da[which(Da$geo_takis == "UA3c" & Da$class == "Fauna"),]$z, bw="nrd"), col="#ff7f00", lwd=2)
abline(v=mean(Da[which(Da$geo_takis == "UA3c" & Da$class == "Fauna"),]$z), col="#ff7f00", lwd=1, lty=2)
### fauna UA4
lines(density(Da[which(Da$geo_takis == "UA4" & Da$class == "Fauna"),]$z, bw="nrd"), col="#ff7f00", lwd=2)
abline(v=mean(Da[which(Da$geo_takis == "UA4" & Da$class == "Fauna"),]$z), col="#ff7f00", lwd=1, lty=2)

legend(x="topright", c("UA3c/UA4 surface","debris/chip","flake","tool","fauna"), lty=1, lwd=2, col=c("grey","#e41a1c", "#377eb8", "#4daf4a", "#ff7f00"))

dev.off()
```

### Area B

```{r}
# subset points until 2015
summary(Xb)
Xb$date <- as.Date(Xb$date, format="%Y-%m-%d")
library(dplyr)
Xb <- dplyr::filter(Xb, date < as.Date("2016-01-01"))

# convert dataframe to ppp, Area B
w <- as.owin(unitsB_2015.shp)
names(Xb)
m <- Xb[,3:14]
Xb.ppp <- ppp(Xb$x, Xb$y, window=w, marks=m)
unitname(Xb.ppp) <- c("meter","meters")
summary(Xb.ppp)
## remove rejected points
Xb.ppp <- as.ppp(Xb.ppp)
## deal with duplicated points
any(duplicated(Xb.ppp))
which(duplicated(Xb.ppp) == "TRUE")
Xb.ppp <- Xb.ppp[!duplicated(Xb.ppp)]

## plot
plot(Xb.ppp$x, Xb.ppp$y, axes=TRUE, asp=1); plot(w, add=TRUE)
plot(Xb.ppp$marks$class, main="Distribution of remains from area B")
plot(droplevels(Xb.ppp$marks$geo_takis), main="Distribution of remains from area A")
```

Xb.ppp
Marked planar point pattern:  1433 points
Average intensity 49.44828 points per square meter
Window area = 29 square meters
UB4c   :1243
UB5a   : 101

```{r}
# extract surface elevation at X positions and calculate minimum distance
summary(Xb.ppp)
S <- UB5a_surf.idw[Xb.ppp]
summary(S)
d <- Xb.ppp$marks$z-S
Db <- cbind(as.data.frame(Xb.ppp), d)
summary(Db)
```

Db: 1433 obs.

#### Fig.8

```{r}
# colorbrewer2.org, 5 classes, qualitative
# col <- c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")
```

```{r}
# fig.8: Vertical distribution area B

setEPS()

postscript("./doc/artwork/Fig8.eps", width=10, height=10)

pdf("./doc/artwork/Fig8.pdf", width=20, height=10)

par(ps=12, cex=1, cex.main=1, cex.axis=1, cex.lab=1)
par(mfcol=c(1,2), ps=12, cex=1, cex.main=1, cex.axis=1, cex.lab=1)

## UB5a
hist(Db[which(Db$geo_takis == "UB5a" & Db$d <= 0),]$d, nclass=5, prob=TRUE, xlim=c(-0.6,0.1), ylim=c(0,14), xlab="d (meters)", main="a")
rug(Db[which(Db$geo_takis == "UB5a" & Db$d <= 0 & Db$type == "debris"),]$d, col="#e41a1c", lwd=2)
lines(density(Db[which(Db$geo_takis == "UB5a" & Db$d <= 0 & Db$type == "debris"),]$d, bw="nrd"), col="#e41a1c", lwd=2)
abline(v=mean(Db[which(Db$geo_takis == "UB5a" & Db$d <= 0 & Db$type == "debris"),]$d), col="#e41a1c", lwd=1, lty=2)
rug(Db[which(Db$geo_takis == "UB5a" & Db$d <= 0 & Db$type == "debitage"),]$d, col="#377eb8", lwd=2)
lines(density(Db[which(Db$geo_takis == "UB5a" & Db$d <= 0 & Db$type == "debitage"),]$d, bw="nrd"), col="#377eb8", lwd=2)
abline(v=mean(Db[which(Db$geo_takis == "UB5a" & Db$d <= 0 & Db$type == "debitage"),]$d), col="#377eb8", lwd=1, lty=2)
rug(Db[which(Db$geo_takis == "UB5a" & Db$d <= 0 & Db$type == "tool"),]$d, col="#4daf4a", lwd=2)
lines(density(Db[which(Db$geo_takis == "UB5a" & Db$d <= 0 & Db$type == "tool"),]$d, bw="nrd"), col="#4daf4a", lwd=2)
abline(v=mean(Db[which(Db$geo_takis == "UB5a" & Db$d <= 0 & Db$type == "tool"),]$d), col="#4daf4a", lwd=1, lty=2)
lines(density(Db[which(Db$geo_takis == "UB5a" & Db$d <= 0 & Db$class == "Fauna"),]$d, bw="nrd"), col="#ff7f00", lwd=2)
abline(v=mean(Db[which(Db$geo_takis == "UB5a" & Db$d <= 0 & Db$class == "Fauna"),]$d), col="#ff7f00", lwd=1, lty=2)
legend(x="topright", c("debris","flake","tool","fossil"), lty=1, lwd=2, col=c("#e41a1c", "#377eb8", "#4daf4a", "#ff7f00"))

## UB4c
hist(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0),]$d, nclass=10, prob=TRUE, xlim=c(-0.1,0.6), ylim=c(0,7), xlab="d (meters)", main="UB4c")
rug(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$type == "debris"),]$d, col="#e41a1c", lwd=2)
lines(density(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$type == "debris"),]$d, bw="nrd"), col="#e41a1c", lwd=2)
abline(v=mean(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$type == "debris"),]$d), col="#e41a1c", lwd=1, lty=2)
rug(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$type == "debitage"),]$d, col="#377eb8", lwd=2)
lines(density(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$type == "debitage"),]$d, bw="nrd"), col="#377eb8", lwd=2)
abline(v=mean(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$type == "debitage"),]$d), col="#377eb8", lwd=1, lty=2)
rug(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$type == "tool"),]$d, col="#4daf4a", lwd=2)
lines(density(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$type == "tool"),]$d, bw="nrd"), col="#4daf4a", lwd=2)
abline(v=mean(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$type == "tool"),]$d), col="#4daf4a", lwd=1, lty=2)
rug(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$type == "core"),]$d, col="#984ea3", lwd=2)
lines(density(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$type == "core"),]$d, bw="nrd"), col="#984ea3", lwd=2)
abline(v=mean(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$type == "core"),]$d), col="#984ea3", lwd=1, lty=2)
lines(density(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$class == "Fauna"),]$d, bw="nrd"), col="#ff7f00", lwd=2)
abline(v=mean(Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$class == "Fauna"),]$d), col="#ff7f00", lwd=1, lty=2)
legend(x="topright", c("debris","flake","tool","core","fossil"), lty=1, lwd=2, col=c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00"))

dev.off()
```

## Tests

```{r}
# test normality assumption for the elephas
foo <- Elephas.ppp$z
boxplot(foo)
m <- mean(foo)
s <- sd(foo)
hist(foo, prob=TRUE)
curve(dnorm(x, mean=m, sd=s), col="red", lwd=2, add=TRUE)
qqnorm(foo); qqline(foo, col="red", lwd=2)
shapiro.test(foo)
ks.test(foo, "pnorm")

# mean & sd elevation of erosional contact UA3c/UA4
summary(UA4_surf)
sd(UA4_surf$marks)

# student's t-test: compare debris/fauna distributions
foo <- Da[which(Da$geo_takis == "UA3c" & Da$type == "debris"),]$z
bar <- Da[which(Da$geo_takis == "UA3c" & Da$class == "Fauna"),]$z
boxplot(foo)
m <- mean(foo)
s <- sd(foo)
hist(foo, prob=TRUE)
curve(dnorm(x, mean=m, sd=s), col="red", lwd=2, add=TRUE)
qqnorm(foo); qqline(foo, col="red", lwd=2)
t.test(foo, bar)

foo <- Elephas.ppp$z
bar <- Da[which(Da$geo_takis == "UA4" & Da$class == "Fauna"),]$z
t.test(bar, foo)

foo <- Db[which(Db$geo_takis == "UB5a" & Db$class == "Lithic"),]$z
bar <- Db[which(Db$geo_takis == "UB5a" & Db$class == "Fauna"),]$z
t.test(bar, foo)

foo <- Db[which(Db$geo_takis == "UB4c" & Db$class == "Lithic"),]$z
bar <- Db[which(Db$geo_takis == "UB4c" & Db$class == "Fauna"),]$z
t.test(bar, foo)

# one-way between groups ANOVA
## data
names(Db)
foo <- Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$class == "Lithic"),c(5,15)]
bar <- Db[which(Db$geo_takis == "UB4c" & Db$d >= 0 & Db$class == "Fauna"),c(5,15)]
baz <- rbind(foo, bar)
baz$type <- factor(baz$type)
levels(baz$type)
levels(baz$type) <- c("core","debitage","debris","indet","fauna","fauna","tool","fauna")
baz <- baz[-which(baz$type == "indet"),]
baz <- baz[-which(baz$type == "core"),]
baz$type <- factor(baz$type)
summary(baz)
str(baz)
## eda
boxplot(d ~ type, data=baz)
## one-way between-groups anova
out <- aov(d ~ type, data=baz)
summary(out)
## assessing test assumptions
plot(out)
plot(out, 1) # graphical test of homogeneity
plot(out, 2) # graphical test of normality
## Tukey Honestly Significant Difference test
TukeyHSD(out)
plot(TukeyHSD(out))
## 
library(gplots)
gplots::plotmeans(d ~ type, data=baz)
##
library(multcomp)
tuk <- multcomp::glht(out, linfct=mcp(type="Tukey"))
plot(multcomp::cld(tuk, level=.05), col="lightgrey")
```

## References
