---
title: "Beyond maps: Patterns of formation processes at the Middle Pleistocene open-air site of Marathousa 1, Megalopolis Basin, Greece"
subtitle: "Supplementary information: Testing the reliability of the clock-based circular data for fabric analysis"
author: "Domenico Giusti"
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: "../doc/manuscript/marathousa_rr.bib"
output: 
  html_document: 
    fig_caption: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages: fabric analysis
library(circular)
library(CircStats)
library(RFOC)
library(plotrix)
library(heR.Misc)
```

In our fabric analysis we compared circular data collected using different methods (namely, the clock and the compass methods).

Here we test the reliability of our data collected with the clock method, using a subset of data for which both methods were used to record the orientation of the item.

## Reading and subsetting data

```{r}
# reproducible research from here...
getwd()
setwd("/home/dncgst/projects/marathousa1_qi")

UB4c_test.fabric <- read.csv("data/UB4c_test.fabric.csv", header=TRUE, sep=",", skip=0)
summary(UB4c_test.fabric)
```

The dataset includes:

Summary: 38 obs. (26 KN, 7 LI, 2 LOST.KN, 2 ZN, 1 HZ)

## Comparing the clock and the compass methods for fabric analysis (UB4c)

```{r}
# mean
deg(circ.mean(rad(UB4c_test.fabric$bearing_clock)))
deg(circ.mean(rad(UB4c_test.fabric$bearing_comp)))
```

Mean direction UB4c clock: 91.65627°
Mean direction UB4c compass: 151.1705°

```{r}
# rose diagrams
heR.Misc::rose(UB4c_test.fabric$bearing_clock, bins=72, col="red", rings=TRUE, labels=TRUE, rscale=2, main="")
heR.Misc::rose(UB4c_test.fabric$bearing_comp, bins=72, col="red", rings=TRUE, labels=TRUE, rscale=2, main="")
```

```{r}
# uniformity tests
r.test(rad(UB4c_test.fabric$bearing_clock))
r.test(rad(UB4c_test.fabric$bearing_comp))
kuiper(rad(UB4c_test.fabric$bearing_clock))
kuiper(rad(UB4c_test.fabric$bearing_comp))
watson(rad(UB4c_test.fabric$bearing_clock))
watson(rad(UB4c_test.fabric$bearing_comp))
rao.spacing(rad(UB4c_test.fabric$bearing_clock))
rao.spacing(rad(UB4c_test.fabric$bearing_comp))
```

Rayleig p-values: 0.05 (bearing_clock) vs 0.10 (bearing_comp)
Kuiper: 1.8963 vs 1.3944
Watson: 0.1937 vs 0.1268
Rao: 255.7895 vs 128.5263

All the tests fail to reject the null hypothesis, but Rao for the clock sample...

### QQ plot

```{r}
## angular QQ-plot: comapre the distributional shape of two samples, but not any differences in their central directions, as both samples are centred around their median directions in the construction of the plot

cdat1 <- circular(UB4c_test.fabric$bearing_clock, units="degrees", zero=pi/2, rotation="clock")
plot(cdat1, stack=TRUE, bins=720, cex=1.2, sep=0.04, shrink=0.95)
arrows.circular(mean(cdat1), y=rho.circular(cdat1), lty=1, lwd=2)
arrows.circular(median(cdat1), y=rho.circular(cdat1), lty=2, lwd=3)

cdat2 <- circular(UB4c_test.fabric$bearing_comp, units="degrees", zero=pi/2, rotation="clock")
plot(cdat2, stack=TRUE, bins=720, cex=1.2, sep=0.04, shrink=0.95)
arrows.circular(mean(cdat2), y=rho.circular(cdat2), lty=1, lwd=2)
arrows.circular(median(cdat2), y=rho.circular(cdat2), lty=2, lwd=3)

cdat1 <- circular(UB4c_test.fabric$bearing_clock*2*pi/360)
cdat2 <- circular(UB4c_test.fabric$bearing_comp*2*pi/360)

TwoSampleQQ(cdat1, cdat2)
```

### Watson's Two-sample Test

```{r}
# test for a common distribution
## Performs Watson's test for homogeneity on two samples of circular data.
watson.two.test(cdat1,cdat2)
watson.two.test(cdat1,cdat2,alpha = 0.05)
```


### Woodcock diagram

```{r}
# Angles are in radians, not degrees!!!
plunge <- rad(UB4c_test.fabric$plunge_clock)
bearing <- rad(UB4c_test.fabric$bearing_clock)
plunge <- rad(UB4c_test.fabric$plunge_comp)
bearing <- rad(UB4c_test.fabric$bearing_comp)

# Woodcock diagram
## step 1) express each direction in the sample as a set of three direction cosines. If the data are in the form of trend and plunge, the formulae are:
### l=cos(plunge)*cos(trend)
l <- c(cos(plunge)*cos(bearing))
### m=cos(plunge)*sin(trend)
m <- c(cos(plunge)*sin(bearing))
### n=sin(plunge)
n <- c(sin(plunge))
## step 2) and 3) compute the products of each direction-cosine set. sum the products over the whole data sample to give the orientation tensor
b <- matrix(c(
  sum(l^2),sum(l*m),sum(l*n),
  sum(m*l),sum(m^2),sum(m*n),
  sum(n*l),sum(n*m),sum(n^2)),
  ncol=3, byrow=TRUE)
## step 4) compute the eigenvectors and eigenvalues of this matrix
### the eigenvectors (v1,v2,v3) of this matrix correspond to three orthogonal 'principal axes' of the data sample. Axis v1 is an estimate of the mean direction of the data; v3 is the pole to any girdle present in the data [@Watson1966]
eigenvectors <- eigen(b)$vectors
dimnames(eigenvectors) <- list(c("v1","v2","v3"),c("l","m","n"))
### The eigenvalues (λ1,λ2,λ3), associated with (v1,v2,v3) respectively, describe the shape of the data sample [@Watson1966; @Mark1974; @Woodcock1973; @Woodcock1977]
eigenvalues <- eigen(b)$values
names(eigenvalues) <- c("λ1","λ2","λ3")
## step 5) compute the eigenvalue ratio, particularly the ratio of the largest to smallest eigenvalue (S1/S3=λ1/λ3)
### since λ1+λ2+λ3=N then normalized eigenvalues Sj=λj/N are related by S1+S2+S3=1. A data cluster has S1>S2~=S3 and a data girdle has S1~=S2>S3
s1 <- eigenvalues["λ1"]/length(plunge)
s2 <- eigenvalues["λ2"]/length(plunge)
s3 <- eigenvalues["λ3"]/length(plunge)

# save UB4c_clock
clock.s1_s2 <- s1/s2
clock.s2_s3 <- s2/s3
clock.s1_s3 <- s1/s3
clock.is <- s3/s1
clock.el <- 1-(s2/s1)
clock.bs <- 1-(clock.is+clock.el)

# save UB4c_comp
comp.s1_s2 <- s1/s2
comp.s2_s3 <- s2/s3
comp.s1_s3 <- s1/s3
comp.is <- s3/s1
comp.el <- 1-(s2/s1)
comp.bs <- 1-(comp.is+comp.el)
```

```{r}
## Woodcock diagram
#setEPS()

#pdf("./doc/artwork/FigSI1.pdf", width=10, height=10)

#par(ps=12, cex=1, cex.main=2, cex.axis=1, cex.lab=1)

plot(log(comp.s2_s3),log(comp.s1_s2),pch=15,xlim=c(0,3),ylim=c(0,3),xlab="ln S2/S3",ylab="ln S1/S2", main="Woodcock diagram")
## shape parameter K=ln(S1/S2)/ln(S2/S3) ranges from 0 (uniaxial girdles) to infinite (uniaxial clusters)
abline(h=0,col="grey",lty=2)
abline(v=0,col="grey",lty=2)
abline(0,1,col="grey",lty=2)
abline(0,2,col="grey",lty=2)
abline(0,.5,col="grey",lty=2)
abline(0,5,col="grey",lty=2)
abline(0,.2,col="grey",lty=2)
text(2.95,2.8,label="K=1",col="grey",lwd=.02)
text(2.95,.1,label="K=0",col="grey",lwd=.02)
text(.2,3,label=expression(paste("K=",infinity)),col="grey",lwd=.02)
text(2.95,.75,label="K=0.2",col="grey",lwd=.02)
text(2.95,1.65,label="K=0.5",col="grey",lwd=.02)
text(.8,3,label="K=5",col="grey",lwd=.02)
text(1.7,3,label="K=2",col="grey",lwd=.02)
## complementary to the shape parameter is the parameter C=ln(S1/S3) which expresses the 'strength' of the preferred orientation in the data sample [@Woodcock1977]. Strongly organized samples have larger C and plot further from the origin of the ratio graph. A perfect uniform distribution of data would have C=0 and would plot at the origin with S1=S2=S3=1/3. Random samples will also plot close to the origin and have low C values, suggesting that this parameter might be used as a test statistic for randomness
abline(3,-1,col="grey",lty=1)
text(1.7,1.5,label="C=3",col="grey",lwd=.02)
##

PlotPTsmooth(deg(UB4c_test.fabric$bearing_clock),deg(UB4c_test.fabric$plunge_clock),x=log(clock.s2_s3),y=log(clock.s1_s2),siz=0.1,IMAGE=TRUE,CONT=TRUE,add=TRUE)
text(.25,.40,label="UB4c (clock)",col="black",lwd=.02)

PlotPTsmooth(deg(UB4c_test.fabric$bearing_comp),deg(UB4c_test.fabric$plunge_comp),x=log(comp.s2_s3),y=log(comp.s1_s2),siz=0.1,IMAGE=TRUE,CONT=TRUE,add=TRUE)
text(.80,.20,label="UB4c (comp)",col="black",lwd=.02)

text(1.45,1.65,label="WEAK <--> STRONG",col="gray",lwd=.03,srt=45,pos=1)
text(2.55,2.63,label="<-- CLUSTERS | GIRDLES -->",col="gray",lwd=.03,srt=-45,pos=1)

#dev.off()
```

### Benn diagram

```{r}
## Benn diagram
#setEPS()

#pdf("./doc/artwork/FigSI2.pdf", width=12, height=10)

#par(ps=12, cex=1, cex.main=2, cex.axis=1, cex.lab=1)

triax.plot(matrix(c(comp.bs,comp.el,comp.is),ncol=3,byrow=2), show.grid=TRUE, pch=16, cex.ticks=.8, cex.axis=1, axis.labels=c("","Elongation Index\n EL=1-(S2/S1)","Isotropy Index\n IS=S3/S1"), tick.labels=list(l=seq(0.1,0.9,by=0.1), r=seq(0.1,0.9,by=0.1), b=NULL), main="Benn diagram")

triax.points(matrix(c(clock.bs,clock.el,clock.is),ncol=3,byrow=2), pch=16, cex=2, col="#a51e37")
text(.42,.53,label="UB4c clock")

triax.points(matrix(c(comp.bs,comp.el,comp.is),ncol=3,byrow=2), pch=16, cex=2, col="#a51e37")
text(.48,.40,label="UB4c comp")

text(0.5,0.9,label="isotropic")
text(-0.1,0,label="planar")
text(1.1,0,label="linear")

#dev.off()
```